---
title: "Airline Safety"
author: "Pranesh Krishnamurthy"
date: "25 January 2018"
output: pdf_document
---

This is a reboot of the initial exploration project (which failed).

Setup
---

```{r libraries}

library(tidyverse)
```

```{r data}

airlines <- read_csv("../Data/airlinesafety.csv", col_names = T)
```


# Data wrangling

The data doesn't seem to be in a form that we can easily work with, so we'll need to wrangle it to make it so.

## Convert to tidy data

Firstly, there are a lot of redundant variables. So, I'll gather them under one column and then focus on separating them, so that I can easily analyse the data.

### Gathering

```{r gathering}

airlines <- airlines %>% 
  gather(incidents_85_99:fatalities_00_14, key = "type", value = "count") 
```

### Replacing the underscores with a dot

```{r replace the dots}

airlines$type <- str_replace(airlines$type, "_(?=[0-9])", "\\.")
```

### Separating type and time-span

```{r separating type and time-span}

airlines <- airlines %>% 
  separate(type, c("type", "time_span"), sep = "\\.")
```

### Fixing parsing issues and recoding variables

Now that I've separated the variables, I can proceed to recode them and you know, sorta do stuff with them.

```{r recoding variables}

airlines <- airlines %>% 
  mutate(time_span = fct_recode(time_span,
                                "1985 - 1999" = "85_99",
                                "2000 - 2014" = "00_14"))
```

Now that this is over, the dataset is finally starting to take shape. To make the data tidy, one last step is necessary. Spreading the type of incidents into columns.

```{r spreading type}

airlines <- airlines %>% 
  spread(type, count)
```

With this, the data has been put in a form that is easy to analyze. We'll do just that now.

```{r writing to disk}

write_csv(airlines, "../Derived/airline_safety_cleaned.csv", col_names = T)
```


